Sequence modeling enables direct inference of germline haplotypes from short read sequencing data

### 2022-08-05 manuscript/manuscript.tex
bq figure

\title{Variant calling with transformers}
\author[1]{Brendan O'Fallon}
\author[1]{Ashini Bolia}
\author[1]{Jacob Durtschi}
\author[1]{Luobin Yang}
\author[1]{Eric Frederickson}
\author[1]{Katherine Noble}
\author[1]{Joshua Coleman}
\author[1]{Hunter Best}

@@

\begin{abstract}
	Detection of germline variants in next-generation sequencing data is an essential element of modern genomics. Variant detection tools typically rely on a variety of statistical algorithms and heuristic techniques to identify variants. Here we describe a new approach that replaces handcrafted statistical methods with a single, end-to-end deep learning model. Our model frames variant detection as a sequence-to-sequence modeling task, akin to language translation, and employs a transformer-based architecture to translate alignment columns into two predicted haplotype sequences. We demonstrate that this method predicts variant genotypes and phase accurately, has the ability to realign complex and ambiguous read mappings to produce accurate haplotype predictions, and uses the local surrounding read context to inform predictions about a given position.

\end{abstract}

vim

@@

Accurate inference of the DNA sequence variants in a sample is a fundamental component of genomics analysis. Tools that aim to identify variants must overcome several challenges to produce an accurate set of variant calls. For instance, the sequenced reads themselves may contain errors in the form of miscalled bases or PCR-induced repeat length errors, or the tools used to align reads to the reference genome may produce incorrect mappings (e.g. Li et al. 2014, Goldfeder et al. 2016). Identifying these errors and differentiating them from real genomic variation is challenging, and modern variant detection tools employ a variety of statistical and heuristic techniques to achieve high precision and sensitivity (De Pristo et al. 2011, Nielson et al. 2011).

All variant discovery tools must address two challenges. First, callers must identify the potential alleles in a region. Then, given a set of potential alleles, they must assess the likelihood of each candidate (or pairs of candidates in the diploid case), to determine which should be included in the caller output. Early callers, such as samtools / mpileup (Li et al. 2009) and the UnifiedGenotyper tool from the Genome Analysis ToolKit (GATK, De Pristo et al. 2011) rely on the read aligner to generate candidate allele candidates, and utilize ad-hoc heuristics and thresholds to determine likely alleles (Nielsen et al. 2011). Later tools (e.g. Poplin et al. 2018, Kim et al. 2018, Cooke et al. 2021) incorporated local re-assembly of reads in each region of interest, resulting in a significant improvement in the sensitivity and precision of variant calls. For instance, the HaplotypeCaller tool identifies candidate haplotypes by constructing de Bruijn graphs from k-mers present in the reads, and assesses haplotype likelihoods with a pair Hidden Markov Model (HMM).

@@

Our model consists of a transformer-based encoder and a simple, fixed-length decoder that produces two output sequences, one for each predicted haplotype. The encoder component is an unmodified transformer with GeLU activations as implemented in PyTorch 1.10 ((Vaswani et al. 2017, Paszke et al. 2019). Input tokens are the collection of bases that align to a given reference position across reads, with some modifications described below. We add an additional fully connected layer prior to the transformer encoders which embeds the encoded basecalls in $d$ dimensions, where $d=12$ for the analyses here. The embedded basecalls are `flattened' along the read dimension, producing an input token with size $dr$ where $r$ is the number of reads ($r=100$ for all experiments proposed here).  Instead of the typical 1-dimensional positional encoding, we employed a 2-dimensional encoding, enabling the model to retain information about which read contained a given base across input tokens.

The decoder consists of a single fully-connected layer followed by two additional fully connected components that each generate a single haplotype. Predictions are discrete probability distributions over the four possible bases at each predicted output position, produced with a softmax function over the raw outputs. The decoder is of fixed length and produces two haplotype sequences equal in length to the number of input tokens $g$. The full model architecture is shown in \ref{fig:architecture}.

@@

The random forest classifier is trained by calling variants on selected regions from autosomes 1-4 totalling approximately 120Mb. True and false positive variant calls were identified with the \textit{vcfeval} tool (Cleary et al. 2015). We used the scikit-learn implementation of the random forest with 100 trees, a maximum tree depth of 25, and 'balanced' class weighting option.

@@

>>>>

\title{Sequence modeling enables direct inference of germline haplotypes from short read sequencing data}
\author[1]{Brendan O'Fallon}
\author[1]{Ashini Bolia}
\author[1]{Jacob Durtschi}
\author[1]{Luobin Yang}
\author[1]{Eric Frederickson}
\author[1]{Katherine Noble}
\author[1]{Joshua Coleman}
\author[1]{Hunter Best}

@@

\begin{abstract}
	Detection of germline variants in next-generation sequencing data is an essential component of modern genomics analysis. Variant detection tools typically rely on statistical algorithms such as de Bruijn graphs, Hidden Markov Models and regression models, often coupled with heuristic techniques and thresholds to identify variants. Here we describe a new approach that replaces these handcrafted statistical methods with a single, end-to-end deep learning that directly infers germline haplotypes from short read pileups. Our model frames variant detection as a sequence-to-sequence modeling task, akin to language translation, and employs a transformer-based architecture to translate alignment columns into two predicted haplotype sequences. After training with 17 whole genome sequences from Genome-in-a-Bottle cell lines, we demonstrate that this method predicts variant genotypes and phase accurately, has the ability to realign complex and ambiguous read mappings to produce accurate haplotype predictions, and uses the local read context to inform predictions about a given position. We also demonstrate that a 2-dimensional positional encoding significantly improved precision of the detected variants. Compared to other callers, sensitivity and precision is higher than GATK HaplotypeCaller, but lower than DeepVariant and Strelka2. 
\end{abstract}

@@

Accurate inference of the DNA sequence variants in a sample is a fundamental element of genomics analysis. Tools that aim to identify variants must overcome several challenges to produce an accurate set of variant calls. For instance, the sequenced reads themselves may contain errors in the form of miscalled bases or PCR-induced repeat length errors, or the tools used to align reads to the reference genome may produce incorrect mappings (e.g. Li et al. 2014, Goldfeder et al. 2016).  In addition, large or complex sequence variants may be obscured by limitations in the read alignment algorithm, and require a sophisticated assembly of reads to identify. Reconstructing real variation and discriminating it from technical error is challenging, and modern variant detection tools employ a variety of statistical and heuristic techniques to achieve high precision and sensitivity (e.g. De Pristo et al. 2011, Nielson et al. 2011).

All variant discovery tools must address two challenges. First, callers must identify the potential alleles in a region. Then, given a set of potential alleles, they must assess the likelihood of each candidate (or pairs of candidates in the diploid case), to determine which should be included in the caller output. Early callers, such as samtools / mpileup (Li et al. 2009) and the UnifiedGenotyper tool from the Genome Analysis ToolKit (GATK, De Pristo et al. 2011) rely on the read aligner to generate candidate allele candidates, and utilize several ad-hoc heuristics and thresholds to determine likely alleles (e.g. Nielsen et al. 2011). Later tools (e.g. Poplin et al. 2018, Kim et al. 2018, Cooke et al. 2021) incorporated local re-assembly of reads in each region of interest, resulting in a significant improvement in the sensitivity and precision of variant calls. For instance, the HaplotypeCaller tool identifies candidate haplotypes by constructing de Bruijn graphs from k-mers present in the reads, and assesses haplotype likelihoods with a pair Hidden Markov Model (HMM).

@@

Our model consists of a transformer-based encoder and a simple, fixed-length decoder that produces two output sequences, one for each predicted haplotype. The encoder component is an unmodified transformer with GeLU activations as implemented in PyTorch 1.10 (Vaswani et al. 2017, Paszke et al. 2019). Input tokens are the collection of bases that align to a given reference position across reads, with some modifications described below. We add an additional fully connected layer prior to the transformer encoders which embeds the encoded basecalls in $d$ dimensions, where $d=12$ for the analyses here. The embedded basecalls are `flattened' along the read dimension, producing an input token with size $dr$ where $r$ is the number of reads ($r=100$ for all experiments proposed here).  Instead of the typical 1-dimensional positional encoding, we employed a 2-dimensional encoding, allowing the model to differentiate across both the token (position) and read dimensions. The 2D encoding implementation follows Wang \& Jyh-Charn (2021).

The decoder consists of a single fully-connected layer with ReLU activation followed by two additional fully connected components that each generate a single haplotype. Predictions are discrete probability distributions over the four possible bases at each predicted output position, produced with a softmax function over the raw outputs. The decoder is of fixed length and produces two haplotype sequences equal in length to the number of input tokens $g$. The full model architecture is shown in \ref{fig:architecture}.

@@

The random forest classifier is trained by calling variants on selected regions from autosomes 1-4 totalling approximately 120Mb. True and false positive variant calls were identified with the \textit{vcfeval} tool (Cleary et al. 2015). We used the scikit-learn implementation of the random forest with 100 trees, a maximum tree depth of 25, and `balanced' class weighting option.

\begin{center}
	\begin{tabular}{ l }
	 \textbf{Feature} \\
	 Mean probability of base call at all positions affected by variant \\ 
	 Number of reference bases affected \\
	 Number of alternate bases affected \\
	 Minimum probability of base call at all positions affected by variant \\
	 Maximum probability of base call at all positions affected by variant \\
	 Number of variants in \textit{cis} (same haplotype) with this variant \\
	 Number of variants in \textit{trans} (other haplotype) with this variant \\ 
	 Total number of variants in both \textit{cis} and \textit{trans} \\
	 Distance in bases of call from beginning of window - minimum across windows \\
	 Distance in bases of call from beginning of window - maximum across windows \\
	 Read depth at variant position \\
	 Gnomad genome frequency (v2.0.2) of variant \\
	 Zygosity (heterozygous or homozygous) of variant \\
	 Fraction of reads in BAM/CRAM supporting variant call \\
	 Fraction of reads with mapping quality > 10 supporting variant call \\
	 Total number of reads supporting variant call \\
	\end{tabular}
	\label{table:features}
\end{center}

@@

\vspace{8pt}
Wang, Zelun, and Jyh-Charn Liu. "Translating math formula images to LaTeX sequences using deep neural networks with sequence-level training." International Journal on Document Analysis and Recognition (IJDAR) 24.1 (2021): 63-75.
